---
title: "**Exploratory analysis of the Methodological Quality and Reporting of GLMM in Sports: a Scoping Review**"
output: pdf_document
date: ""
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(readxl)
library(ggplot2)
library(dplyr)
library(patchwork)
library(descr)
library(kableExtra)
library(class)
library(lessR)
library(forcats)
library(stringr)
library(tidyr)
library(compareGroups)
library(camcorder)

theme_set(theme_bw())
```

```{r}
bd <- read_excel("../Data/sr_GLMM_results_data.xlsx")
names(bd) <- str_replace_all(names(bd), ":", "")
bd <- bd |> mutate(across(is.character, ~dplyr::recode(.x, NR = "Not Reported", .default = .x)))
bd <- dplyr::rename(bd, all_of(c("Participants category" = "Category participants")))
```

# General characteristics of the selected articles

```{r}
# summary(bd)
```


```{r, eval = F, results = "hide"}
tab <- descrTable(~  Country + `Journal Name` + `Type of Journal` +  `N (participants)` + `Age (participants)` + `Outcome type` + `Type of design` + `Multilevel (nested design)`,
                  data = bd,
                  include.miss = TRUE, show.n = FALSE)
export2md(tab, include.miss = TRUE, nmax = FALSE,
          caption = "General characteristics of the GLMM models of the N = 55 selected articles in the scoping review, frequencies (percentages) for categorical variables and median (IQR) for continuous variables.",
          header.labels = c("all" = "N (%)"))
```



```{r, fig.width = 20*0.393701, fig.height = 26*0.393701, fig.dpi = 300, fig.cap = "Distribution of the countries where the data from the articles come from (panel (A)) and the name of the journal where the articles were published (panel (B)) among the 55 selected articles."}
p1 <- bd |> 
  mutate(Country = case_when(!(Country %in% c("USA", "International", "Australia", "Spain",
                                              "Canada", "Netherlands", "Norway", "UK")) ~ "Others",
                             .default = Country)) |> 
  ggplot(aes(x = fct_relevel(fct_rev(fct_infreq(Country)), "Others", after = 0))) +
  geom_bar(fill = "skyblue") +
  geom_text(stat = "count", aes(label = paste("", after_stat(count), " (", round(after_stat(count)/sum(after_stat(count))*100, 2), "%)")), hjust = 0, size = 4, nudge_y = 0.2) +
  coord_flip() +
  labs(title = "Country", y = "N (number of articles)", x = "") +
   scale_y_continuous(expand = expansion(mult = c(0.03, 0.35))) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.4)),
        axis.text.y = element_text(size = rel(1.3)),
        axis.title = element_text(size = rel(1.4)))

p2 <- bd |> 
  mutate(`Journal Name` = case_when(!(`Journal Name` %in% c("Journal of Quantitative Analysis in Sports",
                                              "Frontiers in Psychology",
                                              "Frontiers in Sports and Active Living",
                                              "Journal of Sports Science and Medicine",
                                              "PLoS One",
                                              "Journal of Applied Statistics",
                                              "Scientific Reports")) ~ "Others",
                             .default = `Journal Name`)) |> 
  ggplot(aes(x = fct_relevel(fct_rev(fct_infreq(`Journal Name`)), "Others", after = 0))) +
  geom_bar(fill = "skyblue") +
  geom_text(stat = "count", aes(label = paste("", after_stat(count), " (", round(after_stat(count)/sum(after_stat(count))*100, 2), "%)")), hjust = 0, size = 4, nudge_y = 0.2) +
  coord_flip() +
  labs(title = "Journal", y = "N (number of articles)", x = "") +
  scale_x_discrete(label = scales::label_wrap(25)) +
   scale_y_continuous(expand = expansion(mult = c(0.03, 0.35))) +
  theme(legend.position = "none",
        plot.title = element_text(size = rel(1.4)),
        axis.text.y = element_text(size = rel(1.3)),
        axis.title = element_text(size = rel(1.4)))

p1 / p2 + 
  plot_layout(guides = "collect", axes = "collect_x") +
  plot_annotation(tag_levels = "A", tag_prefix = "(", tag_suffix = ")")
```


```{r, eval = F, results = "hide"}
tab <- descrTable(~ Sport + Gender+ `Participants category` + `Category classification`,
                  data = bd,
                  include.miss = TRUE, show.n = FALSE)
export2md(tab, include.miss = TRUE, nmax = FALSE,
          caption = "Characteristics of the sport in the N = 55 selected articles in the scoping review, frequencies (percentages) for categorical variables and median (IQR) for continuous variables.",
          header.labels = c("all" = "N"))
```

```{r, eval = F, results = "hide"}
tab <- descrTable(~  `Test for fixed effects`  + `Test for random effects` + `Variance estimates of random effects` + `Statistical software` +  `Estimation method` + `Data shared` + `Code shared` + `Variable response distribution` + `Overdispersion evaluation` + `Method of variable selection` + `Method of model selection` + `GLMM Validation`,
                  data = bd,
                  include.miss = TRUE, show.n = FALSE)
export2md(tab, include.miss = TRUE, nmax = FALSE,
          caption = "General characteristics of the GLMM models of the N = 55 selected articles in the scoping review, frequencies (percentages) for categorical variables and median (IQR) for continuous variables.",
          header.labels = c("all" = "N"))
```


**Overdispersion evaluation** 
(NOTE: The response variables eligible to test for overdispersion are variables with Poisson and multinomial distributions.)
```{r}
# unique(bd$`Variable response distribution`)
bd  |> 
  mutate(need_overdispersion = `Variable response distribution` %in% c("Poisson", "Multinomial", "Poisson, Negative-Binomial")) |> 
  count(need_overdispersion, `Overdispersion evaluation`) |> 
  pivot_wider(names_from = `Overdispersion evaluation`, values_from = n) |> 
  mutate(frac = Yes/(Yes + `Not Reported`)) 
```


```{r}
tab <- descrTable(~.-`Publication Year`, data = bd, include.miss = TRUE, show.n = FALSE)
export2md(tab, 
          include.miss = TRUE, nmax = FALSE,
          caption = "Characteristics of the N = 55 selected articles in the scoping review, frequencies and percentages.",
          header.labels = c("all" = "N (%)"))
```

In general, the quality and reporting of Generalized Linear Mixed Models in sports is \textbf{\textcolor{red}{poor}}:

- 27.3% of the selected articles did not report the type of study design (whether longitudinal or cross-sectional). 

- 18.2% of the articles did not report the sex of the use case participants.

- \textbf{\textcolor{red}{83.6\%}} of the articles did not report any test for fixed effects and \textbf{\textcolor{red}{NO}} articles reported test for random effects. \textbf{\textcolor{red}{72.7\%}} of the articles did not report the variance estimates of random effects. 

- 12.7% of the articles did not report distribution of the response variable.  91% of the articles did not assess overdispersion in their variables, 78.2% did not report the method for the model selection, and 91% did not report that any validation had been performed on their GLMMs.


\clearpage



```{r, fig.width=7, fig.height=3, fig.cap = "Number of papers published among the 55 selected over the years. The blue line represents the temporal trend."}
bd |> 
  summarise(n = n(), .by = `Publication Year`) |> 
  filter(`Publication Year` >= 2011) |> 
  ggplot(aes(x = `Publication Year`, y = n)) + 
  geom_line() + geom_point() + 
  geom_smooth(se = FALSE, size = 0.75, method = "lm") +
  ylab("N (number of articles)") +
  scale_x_continuous(breaks = seq(2011, 2023, by = 1)) +
  scale_y_continuous(limits = c(0, 12.5), breaks = seq(0, 12, 2)) +
  labs(caption = "*The 1994 article was excluded")
```




## Characteristics of the sport and data



```{r, fig.width=23*0.393701, fig.height = 28*0.393701, fig.cap = "Distribution of sports disciplines studied (panel (A)), participant gender (panel (B)), and professional categories among participants (panel (C)) according to the type of sport category in the 55 selected articles.", fig.pos="H", out.extra=""}

bd <- bd |> 
  mutate(Sport2 = case_when(Sport %in% c("Rugby", "Swimming", "Australian football", "Ice hockey", "Voleyball", "Bicycling", "Field hockey", "Fishing", "Formula 1", "Leonese wrestling", "Tennis") ~ "Others", .default = Sport
                            ))

vars <- names(bd)[c(ncol(bd), 14:15)] # Sport, Gender, Participants category
j <- 1

for(i in vars) {
  p_name <- paste0("p", j) 
  assign(p_name, ggplot(bd, aes(x = fct_rev(fct_infreq(.data[[i]])))) +
           geom_bar(aes(fill = `Category classification`)) +
           geom_text(aes(label = paste0(..count.., " (", round(prop.table(stat(count))*100, 2), "%)"), y = ..count..), stat = "count", nudge_y = 0.2, hjust = 0) +
           geom_text(aes(label = stat(count), group = `Category classification`), stat = "count", 
                     fontface = "bold", col = "white", vjust = 0.45, position = position_stack(vjust = 0.5)) + 
           ggtitle(i) +
           coord_flip() +
           scale_y_continuous(expand = expansion(mult = c(0.03, 0.25))) +
           ylab("N (number of articles)") + xlab("") +
           scale_fill_manual(values = rev(c("#DF8F44FF", "#00A1D5FF", "#B24745FF")), name = "") +
           theme(legend.title = element_text(size = rel(1.3)),
                 legend.text = element_text(size = rel(1.3)),
                 axis.text.x = element_text(size = rel(1.2)),
                 axis.text.y = element_text(size = rel(1.3)),
                 axis.title = element_text(size = rel(1.4)),
                 plot.title = element_text(size = rel(1.4)))
  )
  j <- j + 1
}
p1 <- p1 + ggtitle("Sport")
p1 / p2 / p3 + 
  plot_layout(guides = "collect", axes = "collect_x") +
  plot_annotation(tag_levels = "A", tag_prefix = "(", tag_suffix = ")") & theme(legend.position = "bottom",
                                                                                legend.title = element_blank())

```



```{r}
bd <- bd |> 
  mutate(Software = if_else(`Statistical software` %in% 
                              c("Statistica", "MLwiN", "WinBUGS", "Stata", "HLM", "HLM5 statistical software",
                                "Predictive Analytics Software", "Mplus"), "Other", 
                            `Statistical software`),
         Software = if_else(Software == "SAS, SPSS", "SAS", Software)) ### WATCH OUT!!!

```

## Characteristics of the methodology

```{r, fig.width = 23*0.393701, fig.height = 27*0.393701, fig.cap="Distribution of the study designs (panel (A)), use of multilevel models (panel (B)), and statistical software employed (panel (C)) across the 55 selected articles.", fig.pos = "H", out.extra=""}
vars <- names(bd)[c(11, 12, ncol(bd))] # type of design, multilevel, software
j <- 1

for(i in vars) {
  p_name <- paste0("p", j) 
  sub_data <- bd |> 
    summarise(n = n(), .by = .data[[i]]) |>
    mutate(pct = n/sum(n),
           {{i}} := fct_reorder(.data[[i]], pct))
  assign(p_name, ggplot(sub_data, aes(x = .data[[i]], y = n)) +
           geom_bar(stat = "identity", fill = "skyblue") +
           geom_text(aes(label = paste("", n, " (", round(pct*100, 2), "%)")), hjust = 0, size = 4, nudge_y = 0.2) +
           scale_x_discrete(label = scales::label_wrap(20)) +
           scale_y_continuous(expand = expansion(mult = c(0.03, 0.25))) +
           coord_flip() +
           labs(title = i, y = "N (number of articles)", x = "") +
           theme(legend.position = "none",
                 plot.title = element_text(size = rel(1.4)),
                 axis.text.y = element_text(size = rel(1.3)),
                 axis.title = element_text(size = rel(1.4)))
  )
  j <- j + 1
}

p1 / p2 / p3 + 
  plot_layout(guides = "collect", axes = "collect_x") +
  plot_annotation(tag_levels = "A", tag_prefix = "(", tag_suffix = ")")

```



# Main conclusions

- Out of 55 selected articles 32 (58.2%) were published in sports journals, 19 (34.5%) in multidisciplinary journals and 4 (7.28%) in statistics journals. Regarding the multilevel design, 43 (78%) of the selected articles reported to have a multilevel design. Moreover, over sixty percent were longitudinal/repeated measure studies, whereas 9% (5 articles) were cross-sectional studies, and 27% (15 articles) did not report their study design.  

- On the use case sports data, the three most predominant sports were soccer (20%), multidisciplinary sports (16.4%) and baseball (12.7%). Over seventy percent of the articles were devoted to the study of sports performance analysis, 23.6% to health and 5.5% to academic performance. In addition, the use case participants were professionals in the 64% of the selected articles, while in 33% they were amateurs. Only three articles (5.4%) included female participants, 27 articles (49%) included male participants and 10 articles (18.2%) did not report the sex of the participants. 

- The majority of the selected articles did not share their data nor their code, only 9 and 8 articles did so, respectively. 42% of the articles used R open-source software in their analyses.







